---
output: html_document
editor_options: 
  chunk_output_type: console
---

## Plumbing and kitchen facilities

Why this is important [to come]


```{r}
#| label: constants-plumbkitch
#| include: false

tabst <- "Percent of occupied housing units lacking complete plumbing or kitchen facilities"

```

```{r}
#| label: libraries-plumbkitch
#| include: false

source(here::here("r", "libraries.r"))
source(here::here("r", "functions.r"))
source(here::here("r", "constants.r"))

```

```{r}
#| label: data-prep-plumbkitch
#| include: false


# tabshells <- readRDS(here::here("data", "acs", "tabshells_2019.rds"))
tabdict <- readRDS(path(dchas2019, "chasdict.rds"))
glimpse(tabdict)

tmp <- tabdict |> filter(table=="T3", str_detect(vname, "est"))
t3 <- readRDS(path(dchwide, "tab3_enhanced.rds"))
glimpse(t3)
tmp |> 
  filter(coltype %in% c("Total", "Subtotal"),
         is.na(desc3), is.na(desc4), is.na(desc5)) |> 
  filter(str_detect(desc2, "lacking")) |> 
  select(vname, desc1, desc2)


# trimmed slightly      
# vname    desc1   desc2                                                                     
# <chr>    <chr>   <chr>                                                                     
# T3_est1  Total:  NA                                                                        
# T3_est2  Owner   NA                                                                        
# T3_est3  Owner   AND lacking complete plumbing or kitchen facilities                       
# T3_est45 Renter  NA                                                                        
# T3_est46 Renter  AND lacking complete plumbing or kitchen facilities                       

(keepests <- paste0("T3_est", c(1, 2, 3, 45, 46)))

t3plumbkitch <- t3 |> 
  select(nygeotype, stabbr, geoid, affgeoid, shortname, fullname, countyfp, countyname,
         all_of(keepests)) |> 
  rename(allunits=T3_est1,
         owners=T3_est2,
         renters=T3_est45) |> 
  mutate(ownplumbkitch = T3_est3,
         rentplumbkitch= T3_est46,
         allplumbkitch=ownplumbkitch + rentplumbkitch,
         own_pct=ownplumbkitch / owners,
         rent_pct=rentplumbkitch / renters,
         alltenure_pct=allplumbkitch / allunits,
         renter_share=renters / allunits)
glimpse(t3plumbkitch)
tmp <- t3plumbkitch |> filter(nygeotype=="state", stabbr=="NY")
tmp |> pivot_longer(-c(nygeotype:countyname))
count(t3plumbkitch, nygeotype)


# tabdata for all types
# nytype groups: state, region, county, city, town, village
tabdata <- t3plumbkitch |> 
  filter(nygeotype %in%
           c("nation", "state", "region", "county", "city", "town", "village",
             "rgn_cities", "rgn_villages", "rgn_xcityvill")) |>
  group_by(nygeotype) |> 
  mutate(ownrank=row_number(desc(own_pct)),
         rentrank=row_number(desc(rent_pct)),
         alltenurerank=row_number(desc(alltenure_pct)),
         shortname=ifelse(nygeotype=="state" & stabbr=="NY",
                          paste0(shortname, " State"),
                          shortname))  |> 
  select(nygeotype, stabbr, shortname, countyname, 
         allunits,
         own_pct, rent_pct, alltenure_pct, 
         ownrank, rentrank, alltenurerank,
         renter_share) |> 
  ungroup() |> 
  arrange(desc(alltenure_pct))

saveRDS(tabdata, here::here("report", "results", "plumbkitch_tabdata.rds")) # for appendix

```


### Housing Lacking Complete Plumbing and Kitchen Facilities in New York and Other States

```{r}
#| label: state-plumbkitch-data
#| include: false

tabdata_st <- tabdata |> 
  filter(nygeotype %in% c("nation", "state"), !stabbr %in% c("DC", "PR")) |> 
  mutate(across(contains("rank"), ~ ifelse(nygeotype=="state", .x, NA_real_)),
         shortname=ifelse(nygeotype=="state", 
                          str_remove(shortname, " State"),
                          shortname))

saveRDS(tabdata_st, here::here("report", "results", "plumbkitch_tabdata_st.rds")) # for appendix

```


New York has the fourth highest cost burden among the states, measured by the percentage of households with costs greater than 30 percent of income. New York ranks #4 rank for both owners and renters. As is true in every state, the share of New York's renters who are cost burdened is far greater than the share of owners.

```{r}
#| label: state-plumbkitch-table
#| include: true
#| echo: false

tt <- "Top 10 states ranked by housing lacking complete plumbing and kitchen facilities, plus U.S. and New York"

tabdata_topn <- tabdata_st |> 
  arrange(desc(alltenure_pct)) |> 
  filter(nygeotype=="nation" | row_number() <= 10 | stabbr=="NY") 

tab <- f_tabranks(tabdata_topn, stubvar="shortname", stubhead="State",
           tabtitle=tt, tabsubtitle=tabst)

tabfile <-  path(tabdir, "plumbkitch_states.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

<!-- {{< include C:/Users/donbo/Documents/R_projects/OSC_Housing/report/_gt_snippet.qmd >}} -->

{{< include ../../_gt_snippet.qmd >}}

### Housing Lacking Complete Plumbing and Kitchen Facilities in New York's Regions

```{r}
#| label: region-plumbkitch-data
#| include: false

tabdata_rgn <- tabdata |> 
  filter(stabbr=="NY", nygeotype %in% c("state", "region")) |>
  mutate(across(contains("rank"), ~ ifelse(nygeotype=="state", NA_real_, .x)))

tabdata_rgn

saveRDS(tabdata_rgn, here::here("report", "results", "plumbkitch_tabdata_rgn.rds")) # for appendix

```

High housing cost burden is far more prevalent in New York City, the Mid-Hudson region, and Long Island than in upstate New York. In all regions, high cost burden for renters is more prevalent than it is for homeowners; this is particularly true in upstate regions.

```{r}
#| label: region-plumbkitch-table
#| include: true
#| echo: false

tt <- "New York regions ranked by housing plumbkitching"
tab <- f_tabranks(tabdata_rgn, stubvar="shortname", stubhead="Region",
         tabtitle=tt)

tabfile <-  path(tabdir, "plumbkitch_regions.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

{{< include ../../_gt_snippet.qmd >}}


#### Within-Region Housing Lacking Complete Plumbing and Kitchen Facilities for City and Village vs. Other areas

```{r}
#| label: regional-breakdown-data-plumbkitch
#| include: false

tabdata_rgn_brkdown <- tabdata |> 
  filter(nygeotype %in% c("region", "rgn_cities", "rgn_villages", "rgn_xcityvill")) |> 
  select(nygeotype, shortname, alltenure_pct) |> 
  pivot_wider(names_from = nygeotype, values_from = alltenure_pct) |> 
  select(shortname, region, rgn_cities, rgn_villages, rgn_xcityvill) |> 
  arrange(desc(region))

tabdata_rgn_brkdown

saveRDS(tabdata_rgn_brkdown, here::here("report", "results", "plumbkitch_tabdata_rgn_brkdown.rds")) # for appendix


```


```{r}
#| label: regional-breakdown-table-plumbkitch
#| include: true
#| echo: false

# region rgn_cities rgn_villages rgn_xcityvill
# New York regions ranked by housing cost burden
# Percent of occupied housing units with housing costs greater than 30% of income

tab <- tabdata_rgn_brkdown |> 
  gt() |> 
  sub_missing(columns = everything()) |> 
  tab_header(
      title = "Breakdown of housing lacking complete plumbing and kitchen facilities within New York regions",
      subtitle=tabst) |> 
  cols_label(shortname="Region",
             region=html("Region total"),
             rgn_cities=html("Cities"),
             rgn_villages=html("Villages"),
             rgn_xcityvill=html("Remainder of region")) |> 
  tab_spanner(columns = c(rgn_cities, rgn_villages, rgn_xcityvill),
              label=html("Prevalence within region")) |> 
  fmt_percent(columns=c(region, rgn_cities, rgn_villages, rgn_xcityvill), decimals=1) |> 
  tab_source_note(source_note = "Source: HUD 5-year CHAS ending 2019")
  
tabfile <-  path(tabdir, "plumbkitch_region_breakdown.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

# f_tabranks

```

{{< include ../../_gt_snippet.qmd >}}


### Housing Lacking Complete Plumbing and Kitchen Facilities in New York's Counties

```{r}
#| label: county-plumbkitch-data
#| include: false

tabdata_cnty <- tabdata |> 
  filter(stabbr=="NY", nygeotype %in% c("state", "county")) |>  
  mutate(across(contains("rank"), ~ ifelse(nygeotype=="state", NA_real_, .x)))

tabdata_cnty
saveRDS(tabdata_cnty, here::here("report", "results", "plumbkitch_tabdata_cnty.rds")) # for appendix

```

```{r}
#| label: county-plumbkitch-table
#| include: true
#| echo: false
tt <- "Top 10 counties in New York with greatest lack of complete plumbing and kitchen facilities"

tabdata_topn <- tabdata_cnty |> 
  arrange(desc(alltenure_pct)) |> 
  filter(nygeotype=="state" | row_number() <= 10) 

tab <- f_tabranks(tabdata_topn, stubvar="shortname", stubhead="County",
         tabtitle=tt, tabsubtitle=tabst)

tabfile <-  path(tabdir, "plumbkitch_counties.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}
```

{{< include ../../_gt_snippet.qmd >}}


### Housing Cost Burden in New York's 6 Largest Cities

```{r}
#| label: city-plumbkitch-data
#| include: false

tabdata_city <- tabdata |> 
  filter(stabbr=="NY", nygeotype %in% c("state", "city")) |>  
  mutate(across(contains("rank"), ~ ifelse(nygeotype=="state", NA_real_, .x)))

tabdata_city

saveRDS(tabdata_city, here::here("report", "results", "plumbkitch_tabdata_city.rds")) # for appendix

```


```{r}
#| label: city-plumbkitch-table
#| include: true
#| echo: false

tt <- "Largest cities in New York ranked by lack of complete plumbing and kitchen facilities"

# get nyc and brysa
bignames <- c("New York City", "Buffalo", "Rochester", "Yonkers", "Syracuse", "Albany")
tabdata_bigcities <- tabdata_city |> 
  filter(nygeotype=="state" |
           (nygeotype=="city" & shortname %in% bignames)) |> 
  arrange(desc(alltenure_pct))

tab <- f_tabranks(tabdata_bigcities, stubvar="shortname", stubhead="City",
         tabtitle=tt, tabsubtitle=tabst, keepcounty = TRUE)

tabfile <-  path(tabdir, "plumbkitch_cities.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}
```

{{< include ../../_gt_snippet.qmd >}}


