---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Owners and Renters

<!-- #| echo: false -->
<!-- #| message: false -->
<!-- #| warning: false -->
<!-- #| note: false -->

```{r}
#| label: libraries_and_load
#| include: false

# libraries ---------------------------------------------------------------
source(here::here("r", "libraries.r"))
source(here::here("r", "libraries_maps.r"))
source(here::here("r", "functions_maps.r"))

# locations --------------------------------------------------------------
# acsdir <- here::here("data", "acs")


# get data ----------------------------------------------------------------
nycos_shape <- readRDS(here::here("data", "maps", "nycos_shape.rds"))
acsdata <- readRDS(here::here("data", "acs", "acsdata.rds"))
nycentroids <- readRDS(here::here("data", "maps", "nycentroids.rds"))
nygeom <- readRDS(here::here("data", "maps", "nycentroids.rds"))
xwalk <- readRDS(here::here("data", "crosswalks", "nycounty_xwalk.rds"))

```


```{r}
#| label: map_data
#| include: false


# get owner percentages ----------------------------------------
count(acsdata, type)

df1 <- acsdata |> 
  filter(table=="B25003")

ownpct <- df1 |> 
  mutate(var=factor(line, levels=1:3, labels=c("total", "owner", "renter"))) |> 
  select(type, geoid, area, var, estimate) |> 
  pivot_wider(names_from = var, values_from = estimate) |> 
  mutate(ownpct=owner / total,
         rentpct=renter / total)

# |> left_join(xwalk, by="geoid")
# names(mapdata)
# count(mapdata, area, costate)

```


Housing insecurity tends to be far more prevalent among renters than homeowners, as we will see in later sections. Thus, it's important to know what areas of the state have significant pockets of renters.


## Renter percentages by county

Rental occupancy is far more common in the New York City area than in upstate New York. Looking at county averages tells us very little about rental occupancy upstate.


```{r}
#| label: county_tenure_map
#| fig-width: 12
#| fig-height: 12
#| include: true

mdata <- nycos_shape |>
  left_join(ownpct |> select(-area), by="geoid")
  
p1 <- mdata |> 
  ggplot() +
  geom_sf(aes(fill=rentpct)) +
  scale_fill_distiller(palette = "BrBG", 
                       direction=1, 
                       breaks=seq(0, 1, .1), 
                       labels = scales::percent_format(accuracy = 1)) +
  theme_map() + 
  ggtitle("Renter households as percentage of occupied housing") + 
  theme(plot.title=element_text(size=30)) +
  labs(fill="Renter %") +
  geom_text(aes(xlabel, ylabel, label = area), 
             size = 3, fontface = "bold", colour="blue") + # default 0.25
  theme(legend.position = c(0.9, 0.45)) +
  theme(legend.key.size = unit(1, 'cm'), #change legend key size
        legend.key.height = unit(1, 'cm'), #change legend key height
        legend.key.width = unit(1, 'cm'), #change legend key width
        legend.title = element_text(size=14), #change legend title font size
        legend.text = element_text(size=10))

p1



# p1 <- nycomap(mapdata, mapvar="rentpct", maptitle="Renter percentage by county", direction = 1)
# p1
# ggsave(here::here("report", "results", "owners_hhpct.png"), p, width = 10, height = 6, scale=1.5)

```


## Renter percentages, selected cities

However, there are significant pockets of rental housing upstate, particularly in some cities.

```{r}
#| label: cities_tenure_map
#| eval: true
#| fig-width: 12
#| fig-height: 12
#| out-width: 12in
#| include: true

bubble_data <- nycentroids |> 
  filter(atype=="cousub", str_detect(area, "city")) |> 
  arrange(desc(pop5yr)) |> 
  filter(row_number() <= 15) |> 
  left_join(ownpct |> select(geoid, ownpct, rentpct), by="geoid") |> 
  mutate(aname=str_remove(area, " city"))

# mdata <- nycos_shape |>
#     left_join(ownpct |> select(-area), by="geoid")
  
p2 <- nycos_shape |> 
  ggplot() +
  geom_sf(fill="grey99") +
  geom_point(aes(x=loncenter, y=latcenter, size=pop5yr, colour=rentpct), alpha=0.5, stroke=FALSE, data=bubble_data) +
  scale_size_continuous(range = c(3, 8)) +
  scale_colour_distiller(palette = "BrBG", 
                       direction=1, 
                       breaks=seq(0, 1, .1), 
                       labels = scales::percent_format(accuracy = 1)) +
  geom_text_repel(aes(x=loncenter, y=latcenter, label=aname), size=4, data=bubble_data) + 
  ggtitle("Renter households as percentage of occupied housing",
          subtitle="Largest cities outside New York City") + 
  labs(size="Population", colour="Renter %") +
  theme_map() +
  theme(plot.title=element_text(size=30),
        plot.subtitle=element_text(size=20)) +
  theme(legend.position = c(0.9, 0.45)) +
  theme(legend.key.size = unit(1, 'cm'), #change legend key size
        legend.key.height = unit(1, 'cm'), #change legend key height
        legend.key.width = unit(1, 'cm'), #change legend key width
        legend.title = element_text(size=14), #change legend title font size
        legend.text = element_text(size=10))
p2 

```


```{r}
#| label: counties_cities_tenure_map
#| fig-width: 16
#| fig-height: 16
#| eval: false
#| include: false
# p1 + p2

```


## List by county

```{r}
#| label: county_table

tabdata <- ownpct |> 
  filter(type=="county") |> 
  mutate(aname=str_remove(area, " County, New York")) |> 
  arrange(desc(rentpct))

tab <- tabdata |> 
  select(aname, owner, renter, total, rentpct) |> 
  gt() |> 
  tab_header(
    title = "Owner and renter occupied households in New York State by county",
    subtitle = "Source: 5-year ACS ending 2020"
  ) |> 
  tab_spanner(columns = c(owner, renter, total),
              label="Number of households") |> 
  cols_label(aname="County",
             rentpct = html("Renter households<br>as % of total")) |> 
  fmt_number(columns=c(owner, renter, total),
             decimals=0) |> 
  fmt_percent(columns=rentpct, decimals=1)
tab  

```


## List by city

Note: I dropped Sherrill and have to find it.

```{r}
#| label: city_table

tabdata <- nycentroids |> 
  filter(atype=="cousub", str_detect(area, "city"), geoid!="3609928640") |> # drop Geneva in Fulton
  select(geoid, area, county) |> 
  mutate(area=str_remove(area, " city")) |> 
  left_join(ownpct |> select(-area), by = "geoid") |> 
  left_join(ownpct |> 
              filter(type=="county") |> 
              select(county=area, cntyrentpct=rentpct) |> 
              mutate(county=str_remove(county, " County, New York")),
            by = "county") |> 
  arrange(desc(rentpct))

tab <- tabdata |> 
  select(area, owner, renter, total, rentpct, county, cntyrentpct) |> 
  gt() |> 
  tab_header(
    title = "Owner and renter occupied households in New York State by city (outside NYC)",
    subtitle = "Source: 5-year ACS ending 2020"
  ) |> 
  tab_spanner(columns = c(owner, renter, total),
              label="Number of households in the city") |> 
  tab_spanner(columns = c(county, cntyrentpct),
              label="County name and rental percent") |> 
  cols_label(area="City",
             rentpct = html("Renter households<br>as % of total")) |> 
  fmt_number(columns=c(owner, renter, total),
             decimals=0) |> 
  fmt_percent(columns=c(rentpct, cntyrentpct), decimals=1)
tab  

 
```

