---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Appendix

```{r}
#| label: constants
#| include: false


```

```{r}
#| label: libraries
#| include: false

source(here::here("r", "libraries.r"))
source(here::here("r", "functions.r"))
source(here::here("r", "constants.r"))

```

## Supplemental Housing Cost Burden Tables

```{r}
#| include: false
#| echo: false

# subtitle
tabst <- "Percent of occupied housing units with housing costs greater than 30% of income"
```


### All States

```{r}
#| label: state-cost-table
#| include: true
#| echo: false

tabdata_st <- readRDS(here::here("report", "results", "cb_tabdata_st.rds")) 

tt <- "States ranked by housing cost burden, plus United States"

tab <- f_tabranks(tabdata_st, stubvar="shortname", stubhead="State",
           tabtitle=tt, tabsubtitle=tabst)

tabfile <-  path(tabdir, "cost_states_all.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

{{< include ../_gt_snippet.qmd >}}


### All Counties

```{r}
#| label: county-cost-table
#| include: true
#| echo: false

tabdata_cnty <- readRDS(here::here("report", "results", "cb_tabdata_cnty.rds")) 

tt <- "Counties ranked by housing cost burden, plus New York State"

tab <- f_tabranks(tabdata_cnty, stubvar="shortname", stubhead="County",
           tabtitle=tt, tabsubtitle=tabst)

tabfile <-  path(tabdir, "cost_counties_all.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

{{< include ../_gt_snippet.qmd >}}

### All Cities

Note the high cost burden for Ithaca in Tompkins County. This probably is an artifact of the high student population in Ithaca, where Cornell University and Ithaca College are located.

```{r}
#| label: city-cost-table
#| include: true
#| echo: false

tabdata_city <- readRDS(here::here("report", "results", "cb_tabdata_city.rds")) 

tt <- "Cities ranked by housing cost burden, plus New York State"

tab <- f_tabranks(tabdata_city, stubvar="shortname", stubhead="City",
           tabtitle=tt, tabsubtitle=tabst, keepcounty=TRUE)

tabfile <-  path(tabdir, "cost_cities_all.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

{{< include ../_gt_snippet.qmd >}}


## Supplemental Housing Crowding Tables

```{r}
#| include: false
#| echo: false

# subtitle
tabst <- "Percent of occupied housing units with more than one person per room"

```


### All States

```{r}
#| label: state-crowd-table
#| include: true
#| echo: false

tabdata_st <- readRDS(here::here("report", "results", "crowd_tabdata_st.rds")) 

tt <- "States ranked by housing crowding, plus United States"

tab <- f_tabranks(tabdata_st, stubvar="shortname", stubhead="State",
           tabtitle=tt, tabsubtitle=tabst)

tabfile <-  path(tabdir, "crowd_states_all.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

{{< include ../_gt_snippet.qmd >}}


### All Counties

```{r}
#| label: county-crowd-table
#| include: true
#| echo: false

tabdata_cnty <- readRDS(here::here("report", "results", "crowd_tabdata_cnty.rds")) 

tt <- "Counties ranked by housing crowding, plus New York State"

tab <- f_tabranks(tabdata_cnty, stubvar="shortname", stubhead="County",
           tabtitle=tt, tabsubtitle=tabst)

tabfile <-  path(tabdir, "crowd_counties_all.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

{{< include ../_gt_snippet.qmd >}}

### All Cities

Note that Ithaca in Tompkins County does NOT have high crowding even thought it had high cost burden in the table above. The high cost burden probably is an artifact of the high student population in Ithaca, where Cornell University and Ithaca College are located, reflecting low incomes of college student even though they may have other means of support.

```{r}
#| label: city-crowd-table
#| include: true
#| echo: false

tabdata_city <- readRDS(here::here("report", "results", "crowd_tabdata_city.rds")) 

tt <- "Cities ranked by housing crowding, plus New York State"

tab <- f_tabranks(tabdata_city, stubvar="shortname", stubhead="City",
           tabtitle=tt, tabsubtitle=tabst, keepcounty=TRUE)

tabfile <-  path(tabdir, "crowd_cities_all.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

{{< include ../_gt_snippet.qmd >}}


## Supplemental Tables for Housing Lacking Complete Plumbing and Kitchen Facilities

```{r}
#| include: false
#| echo: false

# subtitle
tabst <- "Percent of occupied housing units lacking complete plumbing or kitchen facilities"

```


### All States

```{r}
#| label: state-plumbkitch-table
#| include: true
#| echo: false

tabdata_st <- readRDS(here::here("report", "results", "plumbkitch_tabdata_st.rds")) 

tt <- "States ranked by lack of complete plumbing and kitchen facilities, plus United States"

tab <- f_tabranks(tabdata_st, stubvar="shortname", stubhead="State",
           tabtitle=tt, tabsubtitle=tabst)

tabfile <-  path(tabdir, "plumbkitch_states_all.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

{{< include ../_gt_snippet.qmd >}}


### All Counties

```{r}
#| label: county-plumbkitch-table
#| include: true
#| echo: false

tabdata_cnty <- readRDS(here::here("report", "results", "plumbkitch_tabdata_cnty.rds")) 

tt <- "Counties ranked by lack of complete plumbing and kitchen facilities, plus New York State"

tab <- f_tabranks(tabdata_cnty, stubvar="shortname", stubhead="County",
           tabtitle=tt, tabsubtitle=tabst)

tabfile <-  path(tabdir, "plumbkitch_counties_all.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

{{< include ../_gt_snippet.qmd >}}

### All Cities

Note that Ithaca in Tompkins County does NOT have high crowding even thought it had high cost burden in the table above. The high cost burden probably is an artifact of the high student population in Ithaca, where Cornell University and Ithaca College are located, reflecting low incomes of college student even though they may have other means of support.

```{r}
#| label: city-plumbkitch-table
#| include: true
#| echo: false

tabdata_city <- readRDS(here::here("report", "results", "plumbkitch_tabdata_city.rds")) 

tt <- "Cities ranked by housing plumbkitching, plus New York State"

tab <- f_tabranks(tabdata_city, stubvar="shortname", stubhead="City",
           tabtitle=tt, tabsubtitle=tabst, keepcounty=TRUE)

tabfile <-  path(tabdir, "plumbkitch_cities_all.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

{{< include ../_gt_snippet.qmd >}}


