
## Housing Cost Burden

<!-- Housing Cost Burden and Its Consequences -->
<!-- import, measures, thresholds -->

### Key points

```{r}
#| label: constants
#| include: false

minunits <- 500
nplaces <- 25

tabst <- "Percent of occupied housing units with housing costs greater than 30% of income"

```


```{r}
#| include: false

source(here::here("r", "libraries.r"))
source(here::here("r", "functions.r"))
source(here::here("r", "constants.r"))

chdir <- here::here("data", "chas")
chdir2019 <- path(chdir, "2015-2019")

# cost <- readRDS(here::here("data", "cost.rds"))

t9 <- readRDS(path(chdir2019, "t9long.rds"))
# we need cnty, cntyname maybe shortname also

# use allrace for this set of tables
cost_wide <- t9 |> 
  filter(race=="allrace") |> 
  # filter(nytype=="state", stabbr=="NY") |> # uncomment for testing
  pivot_wider(names_from = c(tenure, cost)) |> 
  # get own and rent sums where cost burden is computable
  mutate(own_check=own_costle30 + own_costgt30le50 + own_costgt50,
         rent_check=rent_costle30 + rent_costgt30le50 + rent_costgt50,
         own_cost30=own_costgt30le50 + own_costgt50,
         rent_cost30=rent_costgt30le50 + rent_costgt50,
         alltenure_cost30=own_cost30 + rent_cost30,
         own_cost50=own_costgt50,
         rent_cost50=rent_costgt50,
         alltenure_cost50=own_cost50 + rent_cost50) |> 
  relocate(own_check, rent_check, .after=cntyname) |> 
  # calc percentages
  mutate(own_pct30=own_cost30 / own_allcost,
         rent_pct30=rent_cost30 / rent_allcost,
         alltenure_pct30=alltenure_cost30 / alltenure_allcost,
         own_pct50=own_cost50 / own_allcost,
         rent_pct50=rent_cost50 / rent_allcost,
         alltenure_pct50=alltenure_cost50 / alltenure_allcost, 
         renter_share=rent_allcost / alltenure_allcost)

# tabdata for all types
# nytype groups: state, region, county, city, town, village
tabdata <- cost_wide |> 
  filter(nytype %in% c("state", "region", "county", "city", "town", "village")) |>
  rename(allunits=alltenure_allcost,
         own_pct=own_pct30, rent_pct=rent_pct30, alltenure_pct=alltenure_pct30) |> 
  group_by(nytype) |> 
  mutate(ownrank=row_number(desc(own_pct)),
         rentrank=row_number(desc(rent_pct)),
         alltenurerank=row_number(desc(alltenure_pct)))  |> 
  select(nytype, stabbr, mininame, cntyname, 
         allunits,
         own_pct, rent_pct, alltenure_pct, 
         ownrank, rentrank, alltenurerank,
         renter_share) |> 
  ungroup() |> 
  arrange(desc(alltenure_pct))

```


### Housing Cost Burden in New York and Other States

```{r}
#| label: state-cost-data
#| include: false

tabdata_st <- tabdata |> 
  filter(nytype=="state", !stabbr %in% c("DC", "PR")) |> 
  mutate(mininame=str_remove(mininame, " State"))

# tabdata |> 
#   ggplot(aes(own_pct, rent_pct))+
#   geom_point() +
#   geom_text(aes(label=stabbr), hjust=0, vjust=0) +
#   geom_abline(slope=1, intercept=0)


```

New York has the fourth highest cost burden among the states, measured by the percentage of households with costs greater than 30 percent of income. New York ranks #4 rank for both owners and renters. As is true in every state, the share of New York's renters who are cost burdened is far greater than the share of owners.

```{r}
#| label: state-cost-table
#| include: true
#| echo: false

tt <- "States ranked by housing cost burden"

tab <- f_tabranks(tabdata_st, stubvar="mininame", stubhead="State",
           tabtitle=tt, tabsubtitle=tabst)

tabfile <-  path(tabdir, "cost_states.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

<!-- {{< include C:/Users/donbo/Documents/R_projects/OSC_Housing/report/_gt_snippet.qmd >}} -->

{{< include ../../_gt_snippet.qmd >}}


### Housing Cost Burden in New York's Regions

```{r}
#| label: region-cost-data
#| include: false

tabdata_rgn <- tabdata |> 
  filter(stabbr=="NY", nytype %in% c("state", "region")) |>
  mutate(across(contains("rank"), ~ ifelse(nytype=="state", NA_real_, .x)))

tabdata_rgn

```

High housing cost burden is far more prevalent in New York City, the Mid-Hudson region, and Long Island than in upstate New York. In all regions, high cost burden for renters is more prevalent than it is for homeowners; this is particularly true in upstate regions.

```{r}
#| label: region-cost-table
#| include: true
#| echo: false

tt <- "New York regions ranked by housing cost burden"
tab <- f_tabranks(tabdata_rgn, stubvar="mininame", stubhead="Region",
         tabtitle=tt)

tabfile <-  path(tabdir, "cost_regions.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```


{{< include ../../_gt_snippet.qmd >}}

### Housing Cost Burden in New York's Counties

```{r}
#| label: county-cost-data
#| include: false

tabdata_cnty <- tabdata |> 
  filter(stabbr=="NY", nytype %in% c("state", "county")) |>  
  mutate(across(contains("rank"), ~ ifelse(nytype=="state", NA_real_, .x)))

tabdata_cnty

```

```{r}
#| label: county-cost-table
#| include: true
#| echo: false
tt <- "Counties in New York ranked by housing cost burden"
tab <- f_tabranks(tabdata_cnty, stubvar="mininame", stubhead="County",
         tabtitle=tt, tabsubtitle=tabst)

tabfile <-  path(tabdir, "cost_counties.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}
```

{{< include ../../_gt_snippet.qmd >}}

### Housing Cost Burden in New York's Cities

```{r}
#| label: city-cost-data
#| include: false

tabdata_city <- tabdata |> 
  filter(stabbr=="NY", nytype %in% c("state", "city")) |>  
  mutate(across(contains("rank"), ~ ifelse(nytype=="state", NA_real_, .x)))

tabdata_city


```

Note the high cost burden for Ithaca in Tompkins County. This probably is an artifact of the high student population in Ithaca, where Cornell University and Ithaca College are located. *(It might make sense to examine differences by age.)*

```{r}
#| label: city-cost-table
#| include: true
#| echo: false

tt <- "Cities in New York ranked by housing cost burden"
tab <- f_tabranks(tabdata_city, stubvar="mininame", stubhead="City",
         tabtitle=tt, tabsubtitle=tabst, keepcounty = TRUE)

tabfile <-  path(tabdir, "cost_cities.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}
```

{{< include ../../_gt_snippet.qmd >}}

### Housing Cost Burden in New York's Towns

There are more than 500 towns in New York, many of which are quite small. The table below ranks towns with at least 500 occupied housing units by the percentage of households with housing costs that exceed 30 percent of income.

```{r}
#| label: town-cost-data
#| include: false

tabdata_town <- tabdata |> 
  filter(stabbr=="NY", nytype %in% c("state", "town"))  |> 
  filter(allunits >= minunits) |> 
  # rerank dropping excluded units
  group_by(nytype) |> 
  mutate(ownrank=row_number(desc(own_pct)),
         rentrank=row_number(desc(rent_pct)),
         alltenurerank=row_number(desc(alltenure_pct))) |> 
  ungroup() |> 
  filter(row_number() <= nplaces) |>  
  mutate(across(contains("rank"), ~ ifelse(nytype=="state", NA_real_, .x)))

tabdata_town

```

Note that the town of Le Ray in Jefferson County is ranked in #5 overall (owners plus renters) among towns with 500+ housing units but is ranked much further down for owners only and for renters only. *This is not an error.* It is a result of Le Ray's extremely high percentage of renters, which means that the overall weighted value for all households is high because cost-burdened households generally are a much larger share for renters than for owners.

```{r}
#| label: town-cost-table
#| include: true
#| echo: false

tt <- paste0("Towns in New York with ",
             minunits, 
             "+ housing units and greatest housing cost burden")
tab <- f_tabranks(tabdata_town, stubvar="mininame", stubhead="Town",
         tabtitle=tt, tabsubtitle=tabst, keepcounty = TRUE)

tabfile <-  path(tabdir, "cost_towns.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}
```

{{< include ../../_gt_snippet.qmd >}}

### Housing Cost Burden in New York's Villages

There are more than 500 villages in New York, many of which are quite small. The table below ranks villages with at least 500 occupied housing units by the percentage of households with housing costs that exceed 30 percent of income.

```{r}
#| label: village-cost-data
#| include: false

tabdata_village <- tabdata |> 
  filter(stabbr=="NY", nytype %in% c("state", "village"))  |> 
  filter(allunits >= minunits) |>   
  # rerank dropping excluded units
  group_by(nytype) |> 
  mutate(ownrank=row_number(desc(own_pct)),
         rentrank=row_number(desc(rent_pct)),
         alltenurerank=row_number(desc(alltenure_pct))) |> 
  ungroup() |> 
  filter(row_number() <= nplaces) |>  
  mutate(across(contains("rank"), ~ ifelse(nytype=="state", NA_real_, .x)))

tabdata_village

```

```{r}
#| label: village-cost-table
#| include: true
#| echo: false

tt <- paste0("Villages in New York with ",
             minunits, 
             "+ housing units and greatest housing cost burden")
tab <- f_tabranks(tabdata_village, stubvar="mininame", stubhead="Village",
         tabtitle=tt, tabsubtitle=tabst, keepcounty = TRUE)

tabfile <-  path(tabdir, "cost_villages.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}
```

{{< include ../../_gt_snippet.qmd >}}

<!-- ### Who has high housing costs? -->

<!-- \[To come - breakdowns by income and race\] -->
