---
output: html_document
editor_options: 
  chunk_output_type: console
---

## Housing Cost Burden

<!-- Housing Cost Burden and Its Consequences -->

<!-- import, measures, thresholds -->

### Key points

[TO COME]

```{r}
#| label: constants
#| include: false

tabst <- "Percent of occupied housing units with housing costs greater than 30% of income"

```

```{r}
#| label: libraries
#| include: false

source(here::here("r", "libraries.r"))
source(here::here("r", "functions.r"))
source(here::here("r", "constants.r"))

```

```{r}
#| label: data-prep
#| include: false


# tabshells <- readRDS(here::here("data", "acs", "tabshells_2019.rds"))
tabdict <- readRDS(path(dchas2019, "chasdict.rds"))
glimpse(tabdict)

tmp <- tabdict |> filter(table=="T9", str_detect(vname, "est"))
t9 <- readRDS(path(dchwide, "tab9_enhanced.rds"))
glimpse(t9)
# we need cnty, cntyname maybe shortname also

# get all-races cost burden
# T9_est1 hunits
# T9_est2 owner occupied
# T9_est38 renter occupied
tmp |> 
  filter(str_detect(desc1, "Renter")) |> 
  filter(str_detect(desc3, "greater than")) |> 
  select(vname, desc2, desc3) |> 
  mutate(desc2=str_sub(desc2, 20, 40))
# 5, 6, 10, 11, 15, 16, 
(owncb3050vars <- paste0("T9_est", seq(5, 35, 5)))
(owncb50pvars <- paste0("T9_est", seq(6, 36, 5)))
(rentcb3050vars <- paste0("T9_est", seq(41, 71, 5)))
(rentcb50pvars <- paste0("T9_est", seq(42, 72, 5)))

t9cb <- t9 |> 
  mutate(owncb3050 = rowSums(across(all_of(c(owncb3050vars)))),
         owncb50p = rowSums(across(all_of(c(owncb50pvars)))),
         rentcb3050 = rowSums(across(all_of(c(rentcb3050vars)))),
         rentcb50p = rowSums(across(all_of(c(rentcb50pvars)))),
         ) |> 
  relocate(owncb3050, owncb50p, rentcb3050, rentcb50p, .before=T9_est1)
glimpse(t9cb)

t9cb2 <- t9cb |> 
  select(nygeotype, stabbr, geoid, affgeoid, shortname, fullname, countyfp, countyname,
         T9_est1, T9_est2, T9_est38, owncb3050, owncb50p, rentcb3050, rentcb50p) |> 
  mutate(own_cost30=owncb3050 + owncb50p,
         rent_cost30=rentcb3050 + rentcb50p,
         alltenure_cost30=own_cost30 + rent_cost30,
         own_pct30=own_cost30 / T9_est2,
         rent_pct30=rent_cost30 / T9_est38,
         alltenure_pct30=alltenure_cost30 / T9_est1,
         renter_share=T9_est38 / T9_est1)
count(t9cb2, nygeotype)

# tabdata for all types
# nytype groups: state, region, county, city, town, village
tabdata <- t9cb2 |> 
  filter(nygeotype %in%
           c("nation", "state", "region", "county", "city", "town", "village",
             "rgn_cities", "rgn_villages", "rgn_xcityvill")) |>
  rename(allunits=T9_est1,
         own_pct=own_pct30, rent_pct=rent_pct30, alltenure_pct=alltenure_pct30) |> 
  group_by(nygeotype) |> 
  mutate(ownrank=row_number(desc(own_pct)),
         rentrank=row_number(desc(rent_pct)),
         alltenurerank=row_number(desc(alltenure_pct)),
         shortname=ifelse(nygeotype=="state" & stabbr=="NY",
                          paste0(shortname, " State"),
                          shortname))  |> 
  select(nygeotype, stabbr, shortname, countyname, 
         allunits,
         own_pct, rent_pct, alltenure_pct, 
         ownrank, rentrank, alltenurerank,
         renter_share) |> 
  ungroup() |> 
  arrange(desc(alltenure_pct))

saveRDS(tabdata, here::here("report", "results", "cb_tabdata.rds")) # for appendix

```


Cost burden is the ratio of housing costs to household income. For renters, housing cost is gross rent (contract rent plus utilities). For owners, housing cost is "select monthly owner costs", which includes mortgage payment, utilities, association fees, insurance, and real estate taxes.

### Housing Cost Burden in New York and Other States

```{r}
#| label: state-cost-data
#| include: false

tabdata_st <- tabdata |> 
  filter(nygeotype %in% c("nation", "state"), !stabbr %in% c("DC", "PR")) |> 
  mutate(across(contains("rank"), ~ ifelse(nygeotype=="state", .x, NA_real_)),
         shortname=ifelse(nygeotype=="state", 
                          str_remove(shortname, " State"),
                          shortname))

saveRDS(tabdata_st, here::here("report", "results", "cb_tabdata_st.rds")) # for appendix

```


```{r}
#| label: hbar_states
#| eval: false
#| include: false
#| echo: false

p <- tabdata_st |> 
  arrange(desc(alltenure_pct)) |> 
  filter(nygeotype=="nation" | row_number() <= 10) |> 
  select(shortname, alltenure_pct) |>
  ggplot() +
  geom_col(aes(x=alltenure_pct, y=reorder(shortname, alltenure_pct)), fill="blue", width=0.5) +
  scale_x_continuous(name="Cost-burdened households as percent of all households",
                     breaks=seq(0, 1, 0.05),
                     labels=percent_format(accuracy=1)) +
  scale_y_discrete(name=NULL) +
  # scale_fill_manual(values=c("blue", "darkgreen")) +
  ggtitle("Cost-Burdened Households as Percent of All Households, by State") +
  theme_bw() +
  legend_notitle
  
p  

```

New York has the fourth highest cost burden among the states, measured by the percentage of households with costs greater than 30 percent of income. New York ranks #4 rank for both owners and renters. As is true in every state, the share of New York's renters who are cost burdened is far greater than the share of owners.

```{r}
#| label: state-cost-table
#| include: true
#| echo: false

tt <- "Top 10 states ranked by housing cost burden, plus United States"

tabdata_topn <- tabdata_st |> 
  arrange(desc(alltenure_pct)) |> 
  filter(nygeotype=="nation" | row_number() <= 10) 

tab <- f_tabranks(tabdata_topn, stubvar="shortname", stubhead="State",
           tabtitle=tt, tabsubtitle=tabst)

tabfile <-  path(tabdir, "cost_states.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

<!-- {{< include C:/Users/donbo/Documents/R_projects/OSC_Housing/report/_gt_snippet.qmd >}} -->

{{< include ../../_gt_snippet.qmd >}}

### Housing Cost Burden in New York's Regions

```{r}
#| label: region-cost-data
#| include: false

tabdata_rgn <- tabdata |> 
  filter(stabbr=="NY", nygeotype %in% c("state", "region")) |>
  mutate(across(contains("rank"), ~ ifelse(nygeotype=="state", NA_real_, .x)))

tabdata_rgn

saveRDS(tabdata_rgn, here::here("report", "results", "cb_tabdata_rgn.rds")) # for appendix

```

High housing cost burden is far more prevalent in New York City, the Mid-Hudson region, and Long Island than in upstate New York. In all regions, high cost burden for renters is more prevalent than it is for homeowners; this is particularly true in upstate regions.

```{r}
#| label: region-cost-table
#| include: true
#| echo: false

tt <- "New York regions ranked by housing cost burden"
tab <- f_tabranks(tabdata_rgn, stubvar="shortname", stubhead="Region",
         tabtitle=tt)

tabfile <-  path(tabdir, "cost_regions.png")
# tabfile <- "E:/R_projects/OSC_Housing/report/gtables/cost_regions.png"
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

{{< include ../../_gt_snippet.qmd >}}


```{r}
#| label: region-hbar
#| include: true
#| echo: false
names(t9cb2)

p <- t9cb2 |> 
  filter(stabbr=="NY", nygeotype %in% c("region")) |> 
  select(shortname, own_cost30, rent_cost30, alltenure_cost30) |> 
  arrange(desc(alltenure_cost30)) |>
  select(shortname, own_cost30, rent_cost30, alltenure_cost30) |> 
  pivot_longer(-c(shortname, alltenure_cost30)) |> 
  mutate(namef=factor(name, 
                      levels=c("rent_cost30", "own_cost30"),
                      labels=c("renters", "owners"))) |> 
  arrange(namef) |> 
  ggplot() +
  geom_col(aes(x=value, y=reorder(shortname, alltenure_cost30), fill=namef), width=0.5) +
  scale_x_continuous(name="Number of cost-burdened households (thousands)",
                     breaks=seq(0, 2e6, 100e3),
                     labels=number_format(scale=1e-3)) +
  scale_y_discrete(name=NULL) +
  scale_fill_manual(values=c("blue", "darkgreen")) +
  ggtitle("Number of Cost-Burdened Households in New York, by Region") +
  theme_bw() +
  legend_notitle
  
p  

```

#### Within-Region Housing Cost Burden for City and Village vs. Other areas

```{r}
#| label: regional-breakdown-data
#| include: false

tabdata_rgn_brkdown <- tabdata |> 
  filter(nygeotype %in% c("region", "rgn_cities", "rgn_villages", "rgn_xcityvill")) |> 
  select(nygeotype, shortname, alltenure_pct) |> 
  pivot_wider(names_from = nygeotype, values_from = alltenure_pct) |> 
  select(shortname, region, rgn_cities, rgn_villages, rgn_xcityvill) |> 
  arrange(desc(region))

tabdata_rgn_brkdown

saveRDS(tabdata_rgn_brkdown, here::here("report", "results", "cb_tabdata_rgn_brkdown.rds")) # for appendix


```


```{r}
#| label: regional-breakdown-table
#| include: true
#| echo: false

# region rgn_cities rgn_villages rgn_xcityvill
# New York regions ranked by housing cost burden
# Percent of occupied housing units with housing costs greater than 30% of income

tab <- tabdata_rgn_brkdown |> 
  gt() |> 
  sub_missing(columns = everything()) |> 
  tab_header(
      title = "Breakdown of housing cost burden within New York regions",
      subtitle=tabst) |> 
  cols_label(shortname="Region",
             region=html("Region total"),
             rgn_cities=html("Cities"),
             rgn_villages=html("Villages"),
             rgn_xcityvill=html("Remainder of region")) |> 
  tab_spanner(columns = c(rgn_cities, rgn_villages, rgn_xcityvill),
              label=html("Cost burden prevalence within region")) |> 
  fmt_percent(columns=c(region, rgn_cities, rgn_villages, rgn_xcityvill), decimals=1) |> 
  tab_source_note(source_note = "Source: HUD 5-year CHAS ending 2019")
  
tabfile <-  path(tabdir, "cost_region_breakdown.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

# f_tabranks

```

{{< include ../../_gt_snippet.qmd >}}


### Housing Cost Burden in New York's Counties

```{r}
#| label: county-cost-data
#| include: false

tabdata_cnty <- tabdata |> 
  filter(stabbr=="NY", nygeotype %in% c("state", "county")) |>  
  mutate(across(contains("rank"), ~ ifelse(nygeotype=="state", NA_real_, .x)))

tabdata_cnty
saveRDS(tabdata_cnty, here::here("report", "results", "cb_tabdata_cnty.rds")) # for appendix

```

```{r}
#| label: county-cost-table
#| include: true
#| echo: false
tt <- "Top 10 counties in New York with highest housing cost burden"

tabdata_topn <- tabdata_cnty |> 
  arrange(desc(alltenure_pct)) |> 
  filter(nygeotype=="state" | row_number() <= 10) 

tab <- f_tabranks(tabdata_topn, stubvar="shortname", stubhead="County",
         tabtitle=tt, tabsubtitle=tabst)

tabfile <-  path(tabdir, "cost_counties.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}
```

{{< include ../../_gt_snippet.qmd >}}


### Housing Cost Burden in New York's 6 Largest Cities

```{r}
#| label: city-cost-data
#| include: false

tabdata_city <- tabdata |> 
  filter(stabbr=="NY", nygeotype %in% c("state", "city")) |>  
  mutate(across(contains("rank"), ~ ifelse(nygeotype=="state", NA_real_, .x)))

tabdata_city

saveRDS(tabdata_city, here::here("report", "results", "cb_tabdata_city.rds")) # for appendix

```


```{r}
#| label: city-cost-table
#| include: true
#| echo: false

tt <- "Largest cities in New York ranked by housing cost burden"

# get nyc and brysa
bignames <- c("New York City", "Buffalo", "Rochester", "Yonkers", "Syracuse", "Albany")
tabdata_bigcities <- tabdata_city |> 
  filter(nygeotype=="state" |
           (nygeotype=="city" & shortname %in% bignames)) |> 
  arrange(desc(alltenure_pct))

tab <- f_tabranks(tabdata_bigcities, stubvar="shortname", stubhead="City",
         tabtitle=tt, tabsubtitle=tabst, keepcounty = TRUE)

tabfile <-  path(tabdir, "cost_cities.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}
```

{{< include ../../_gt_snippet.qmd >}}


```{r}
#| label: city-hbar
#| eval: false
#| include: false
#| echo: false

# don't include this - using raw numbers, NYC is too big for us to make sense of other cities

p <- t9cb2 |> 
  filter(stabbr=="NY", nygeotype %in% c("city")) |> 
  select(shortname, own_cost30, rent_cost30, alltenure_cost30) |> 
  arrange(desc(alltenure_cost30)) |> 
  filter(row_number() <= 10) |>
  select(shortname, own_cost30, rent_cost30, alltenure_cost30) |> 
  pivot_longer(-c(shortname, alltenure_cost30)) |> 
  mutate(namef=factor(name, 
                      levels=c("rent_cost30", "own_cost30"),
                      labels=c("renters", "owners"))) |> 
  arrange(namef) |> 
  ggplot() +
  geom_col(aes(x=value, y=reorder(shortname, alltenure_cost30), fill=namef), width=0.5) +
  scale_x_continuous(name="Number of cost-burdened households (thousands)",
                     breaks=seq(0, 2e6, 100e3),
                     labels=number_format(scale=1e-3)) +
  scale_y_discrete(name=NULL) +
  scale_fill_manual(values=c("blue", "darkgreen")) +
  ggtitle("Number of Cost-Burdened Households in New York, Top Cities") +
  theme_bw() +
  legend_notitle
  
p  

```



<!-- ### Who has high housing costs? -->

<!-- \[To come - breakdowns by income and race\] -->
