# Housing Cost Burden

```{r}
#| label: libraries
#| include: false

source(here::here("r", "libraries.r"))
source(here::here("r", "functions.r"))
source(here::here("r", "constants.r"))

```

```{r}
#| label: prepare-cost-data
#| include: false
#| eval: false

# prepare all the data for this chapter, other than extracts needed for tables
# and graphs

source(here::here("report", "insecurity", "cost", "cost.r"))

```

```{r}
#| label: constants
#| include: false

tabst <- "Percent of occupied housing units with housing costs greater than 30% of income"
tabcb <- "Percent of occupied housing units with housing costs greater than 30% of income"

```

```{r}
#| label: get-chapter-data
#| include: false

# full table 9 cost burden data
t9cb_full <- readRDS(here::here("report", "results", "t9cb_full.rds"))

tabdata <- readRDS(here::here("report", "results", "cb_tabdata.rds")) # for appendix


```

```{r}
#| label: get-text-constants
#| include: false

# What will we want to say in the text in this chapter?



# number of cost burdened households
nycb_num <- t9cb_full |> 
  filter(stabbr=="NY", nygeotype=="state") |> 
  pull(alltenure_cost30)
nycb_num

nycb_pct <- t9cb_full |> 
  filter(stabbr=="NY", nygeotype=="state") |> 
  pull(alltenure_pct30)


# number(1234567, big.mark = ",", scale=1e-6, accuracy=.01)
# scales::label_comma()(1234567)
# scales::label_percent(accuracy=.001)(0.1234567)
# f_pct(0.1234567, accuracy=.01)


```

New York has `r f_comma(nycb_num, scale=1e-6, accuracy=.1)` million cost-burdened households, constituting `r f_pct(nycb_pct, accuracy=.1)` of the state's households.

## Introduction

Analysts commonly define the cost burden of housing as the ratio of housing costs to household income.[^cost-1] Households for which housing costs exceed 30 percent of income are typically considered cost-burdened.

[^cost-1]: For renters, housing cost is gross rent (contract rent plus utilities). For owners, housing cost is "select monthly owner costs", which includes mortgage payment, utilities, association fees, insurance, and real estate taxes.

## Housing cost burden in New York and other states

```{r}
#| label: state-cost-data
#| include: false

tabdata_st <- tabdata |> 
  filter(nygeotype %in% c("nation", "state"), !stabbr %in% c("DC", "PR")) |> 
  mutate(across(contains("rank"), ~ ifelse(nygeotype=="state", rank(.x), NA_real_)),
         .by=nygeotype) |> 
  mutate(shortname=ifelse(nygeotype=="state", 
                          str_remove(shortname, " State"),
                          shortname))

saveRDS(tabdata_st, here::here("report", "results", "cb_tabdata_st.rds")) # for appendix

```

New York has the fourth highest cost burden among the states, measured by the percentage of households with costs greater than 30 percent of income. New York ranks #4 for both owners and renters. As is true in every state, the share of New York's renters who are cost burdened is far greater than the share of owners.

```{r}
#| label: state-cost-table
#| include: true
#| echo: false

tt <- "Top 10 states ranked by housing cost burden, plus United States"

tabdata_topn <- tabdata_st |> 
  arrange(desc(alltenure_pct)) |> 
  filter(nygeotype=="nation" | alltenurerank <= 10) 

tab <- f_tabranks(tabdata_topn, stubvar="shortname", stubhead="State",
           tabtitle=tt, tabsubtitle=tabst)

tabfile <-  path(tabdir, "cost_states.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

<!-- {{< include C:/Users/donbo/Documents/R_projects/OSC_Housing/report/_gt_snippet.qmd >}} -->

{{< include ../../_gt_snippet.qmd >}}

## Housing cost burden in New York's regions

```{r}
#| label: region-cost-data
#| include: false
#| echo: false

tabdata_rgn <- tabdata |> 
  filter(stabbr=="NY", nygeotype %in% c("state", "region")) |>
  mutate(across(contains("rank"), ~ ifelse(nygeotype=="state", NA_real_, .x)))

tabdata_rgn

saveRDS(tabdata_rgn, here::here("report", "results", "cb_tabdata_rgn.rds")) # for appendix

```

High housing cost burden is far more prevalent in New York City, the Mid-Hudson region, and Long Island than in upstate New York. In all regions, high cost burden for renters is more prevalent than it is for homeowners; this is particularly true in upstate regions.

```{r}
#| label: region-cost-table
#| include: false
#| echo: false

tt <- "New York regions ranked by housing cost burden"
tab <- f_tabranks(tabdata_rgn, stubvar="shortname", stubhead="Region",
         tabtitle=tt)

tabfile <-  path(tabdir, "cost_regions.png")
# tabfile <- "E:/R_projects/OSC_Housing/report/gtables/cost_regions.png"
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}

```

{{< include ../../_gt_snippet.qmd >}}

```{r}
#| label: region-hbar
#| include: true
#| echo: false
# names(t9cb_full)

p <- t9cb_full |> 
  filter(stabbr=="NY", nygeotype %in% c("region")) |> 
  select(shortname, own_cost30, rent_cost30, alltenure_cost30) |> 
  arrange(desc(alltenure_cost30)) |>
  select(shortname, own_cost30, rent_cost30, alltenure_cost30) |> 
  pivot_longer(-c(shortname, alltenure_cost30)) |> 
  mutate(namef=factor(name, 
                      levels=c("rent_cost30", "own_cost30"),
                      labels=c("renters", "owners"))) |> 
  arrange(namef) |> 
  ggplot() +
  geom_col(aes(x=value, y=reorder(shortname, alltenure_cost30), fill=namef), width=0.5) +
  scale_x_continuous(name="Number of cost-burdened households (thousands)",
                     breaks=seq(0, 2e6, 100e3),
                     labels=number_format(scale=1e-3)) +
  scale_y_discrete(name=NULL) +
  scale_fill_manual(values=c("blue", "green")) +
  ggtitle("Number of Cost-Burdened Households in New York, by Region") +
  theme_bw() +
  legend_notitle
  
p  

ggsave(here::here("report", "results", "region_cost_hbar.png"), p, width = 10, height = 6, scale=1.5)

```

## Housing cost burden in New York's counties

```{r}
#| label: county-cost-data
#| include: false
#| echo: false

tabdata_cnty <- tabdata |> 
  filter(stabbr=="NY", nygeotype %in% c("state", "county")) |>  
  mutate(across(contains("rank"), ~ ifelse(nygeotype=="state", NA_real_, .x)))

tabdata_cnty
saveRDS(tabdata_cnty, here::here("report", "results", "cb_tabdata_cnty.rds")) # for appendix

```

```{r}
#| label: county-cost-table
#| include: true
#| echo: false

tt <- "Top 10 counties in New York with highest housing cost burden"

tabdata_topn <- tabdata_cnty |> 
  arrange(desc(alltenure_pct)) |> 
  filter(nygeotype=="state" | row_number() <= 10) 

tab <- f_tabranks(tabdata_topn, stubvar="shortname", stubhead="County",
         tabtitle=tt, tabsubtitle=tabst)

tabfile <-  path(tabdir, "cost_counties.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}
```

{{< include ../../_gt_snippet.qmd >}}

## Housing cost burden in New York's 6 largest cities

```{r}
#| label: city-cost-data
#| include: false
#| echo: false

tabdata_city <- tabdata |> 
  filter(stabbr=="NY", nygeotype %in% c("state", "city")) |>  
  mutate(across(contains("rank"), ~ ifelse(nygeotype=="state", NA_real_, .x)))

tabdata_city

saveRDS(tabdata_city, here::here("report", "results", "cb_tabdata_city.rds")) # for appendix

```

```{r}
#| label: city-cost-table
#| include: true
#| echo: false

tt <- "Largest cities in New York ranked by housing cost burden"

# get nyc and brysa
bignames <- c("New York City", "Buffalo", "Rochester", "Yonkers", "Syracuse", "Albany")
tabdata_bigcities <- tabdata_city |> 
  filter(nygeotype=="state" |
           (nygeotype=="city" & shortname %in% bignames)) |> 
  arrange(desc(alltenure_pct))

tab <- f_tabranks(tabdata_bigcities, stubvar="shortname", stubhead="City",
         tabtitle=tt, tabsubtitle=tabst, keepcounty = TRUE)

tabfile <-  path(tabdir, "cost_cities.png")
gtsave(tab, tabfile, zoom=2, expand=20)
# {{< include ../../_gt_snippet.qmd >}}
```

{{< include ../../_gt_snippet.qmd >}}

```{r}
#| label: city-hbar
#| eval: false
#| include: false
#| echo: false

# don't include this - using raw numbers, NYC is too big for us to make sense of other cities

p <- t9cb_full |> 
  filter(stabbr=="NY", nygeotype %in% c("city")) |> 
  select(shortname, own_cost30, rent_cost30, alltenure_cost30) |> 
  arrange(desc(alltenure_cost30)) |> 
  filter(row_number() <= 10) |>
  select(shortname, own_cost30, rent_cost30, alltenure_cost30) |> 
  pivot_longer(-c(shortname, alltenure_cost30)) |> 
  mutate(namef=factor(name, 
                      levels=c("rent_cost30", "own_cost30"),
                      labels=c("renters", "owners"))) |> 
  arrange(namef) |> 
  ggplot() +
  geom_col(aes(x=value, y=reorder(shortname, alltenure_cost30), fill=namef), width=0.5) +
  scale_x_continuous(name="Number of cost-burdened households (thousands)",
                     breaks=seq(0, 2e6, 100e3),
                     labels=number_format(scale=1e-3)) +
  scale_y_discrete(name=NULL) +
  scale_fill_manual(values=c("blue", "darkgreen")) +
  ggtitle("Number of Cost-Burdened Households in New York, Top Cities") +
  theme_bw() +
  legend_notitle
  
p  

```

## Housing cost burden in urbanized areas compared to elsewhere

Five of New York's six largest cities have cost burden that is above the state average. This is generally true of urbanized areas - cities and villages - throughout the state. In every region, high cost burden is more common in cities and villages than in unincorporated areas. The difference between urbanized and unincorporated areas is greater north of New York City, where cities and villages are far more urbanized than outlying areas, than it is on Long Island where towns tend to be highly urbanized. (See the appendix for data.)

## Upstate and downstate differences in housing cost burden

```{r}
#| label: updowndiffs
#| eval: true
#| include: false
#| echo: false

# see E:/R_projects/OSC_Housing/data/acs/decompose_acs_costs_income.r
costinc <- readRDS(here::here("data", "acs", "cost_vs_income.rds"))
# xwalk <- readRDS(here::here("data", "xwalks", "nycounty_xwalk.rds"))
xwalk <- readRDS(here::here("data", "xwalks", "acs_geocodes.rds")) |> 
  filter(stabbr=="NY", nygeotype=="county")

count(costinc, vname)



costinc |> 
  filter(vname=="cost_renter", geotype != "state") |> 
  left_join(xwalk |> select(geoid, shortname, rgn_code, rgn_osc)) |> 
  summarise(n=n(), index=median(index), .by=c(rgn_code, rgn_osc)) |> 
  arrange(desc(index))

costinc |> 
  filter(vname=="cost_renter", geotype != "state") |> 
  left_join(xwalk |> select(geoid, shortname, rgn_code, rgn_osc)) |> 
  filter(!rgn_code %in% c("li", "nyc", "hudson")) |> 
  mutate(mdn=median(index)) |> 
  arrange(desc(index))
  
costinc |> 
  filter(vname=="mhhi_renter", geotype != "state") |> 
  left_join(xwalk |> select(geoid, shortname, rgn_code, rgn_osc)) |> 
  filter(!rgn_code %in% c("li", "nyc", "hudson")) |> 
  mutate(mdn=median(index)) |> 
  arrange(desc(index))
  
costinc |> 
  select(-variable, -estimate) |> 
  pivot_wider(names_from = vname, values_from = index) |>
  left_join(xwalk |> select(geoid, shortname, rgn_code, rgn_osc)) |> 
  mutate(cb=cost_renter / mhhi_renter,
         updown=ifelse(rgn_code %in% c("li", "nyc", "hudson"), "upstate", "downstate"))
  
  
```

The differences between upstate and downstate New York in housing costs, household income, and housing cost burden are substantial. Every one of the 48 counties outside of New York City, Long Island, and the Mid-Hudson region has median rental housing costs that are below the statewide median -- 39 percent below in the typical upstate county. Incomes are lower in all but one of these upstate counties -- about 28 percent lower in the the typical upstate county. With housing costs much lower in upstate New York than downstate, housing cost burdens upstate may often be driven by low incomes rather than high housing costs. (See the appendix for further discussion of this issue and discussion of selected counties.)


## Trends in housing cost burden

Far greater percentages of renters have high housing costs, relative to income, than do homeowners. Furthermore, owner cost burdens fell substantially during the last decade, in part a result of the fall in interest rates, which lowered mortgage costs. By contrast, renter cost burdens were far more stable over most of the last decade.

Both owner cost burden and renter cost burden rose sharply in 2021, reflecting the rise inflation and, for owner cost burden, the rise in mortgage interest rates.

The share of both owner and renter New York households with housing costs has consistently been greater than the national share.

```{r}
#| label: setup-trends
#| include: false

source(here::here("r", "libraries.r"))
source(here::here("r", "functions.r"))
source(here::here("r", "constants.r"))

dacs <- here::here("data", "acs")

tabshells <- readRDS(path(dacs, "tabshells_2019.rds"))
tmp <- tabshells |> select(table, tabname, universe) |> distinct()

# This chapter relies on time-series data from the annual ACS constructed in data_acs.R ----

```

```{r}
#| label: get-rent-burden-trends
#| include: false

# rent burden
vtab("B25070")

b25070ts <- readRDS(here::here("data", "acs", "B25070_timeseries.rds"))

# note the a and b versions of cost burden, where denominators differ
#  a subtracts noncomputable from denominator, b does not
rentcb <- b25070ts |>
  select(geoid, geotype, name, endyear, variable, estimate) |> 
  pivot_wider(names_from = variable, values_from = estimate) |> 
  mutate(rentcb30a=(B25070_007 + B25070_008 + B25070_009 + B25070_010) /
           (B25070_001 - B25070_011),
         rentcb30b=(B25070_007 + B25070_008 + B25070_009 + B25070_010) / B25070_001)

```

```{r}
#| label: get-owner-mortgage-burden-trends
#| include: false

vtab("B25091")
b25091ts <- readRDS(here::here("data", "acs", "B25091_timeseries.rds"))
glimpse(b25091ts)

# a and b versions have denominators as in renters, and are for housing with mortgage
# the all version is all owner households, not just mortgages
owncb <- b25091ts |>
  select(geoid, geotype, name, endyear, variable, estimate) |> 
  pivot_wider(names_from = variable, values_from = estimate) |> 
  mutate(ownmort=B25091_008 + B25091_009 + B25091_010 + B25091_011,
         ownxmort=B25091_019 + B25091_020 + B25091_021 + B25091_022,
         ownmcb30a=ownmort / (B25091_002 - B25091_012),
         ownmcb30b=ownmort / B25091_002,
         owncb30all=(ownmort + ownxmort) / B25091_001)
glimpse(owncb)

owncb |> filter(geotype=="state" & geoid=="36") |> 
  select(name, endyear, B25091_001, B25091_002, B25091_013, owncb30all, ownmcb30b, ownmcb30a)

```

```{r}
#| label: plot-trends
#| fig-width: 12
#| fig-height: 8
#| fig-column: page-right
#| include: true

# use the "b" version of the denominator, to be consistent with CHAS
pdata <- bind_rows(rentcb |> 
                  select(geoid, geotype, name, endyear, value=rentcb30b) |> 
                  mutate(tenure="renter"),
                owncb |> 
                  select(geoid, geotype, name, endyear, value=owncb30all) |> 
                  mutate(tenure="owner")) |> 
  mutate(geotypef=factor(geotype, 
                         levels=c("nation", "state"),
                         labels=c("United States", "New York")),
         tenuref=factor(tenure, 
                        levels=c("owner", "renter"),
                        labels=c("Owners", "Renters")))

capt1 <- "Note: Data from the 2020 annual ACS excluded due to low response rates."
capt2 <- "Source: American Community Survey, Annual, Tables B25070 and B25091."
capt <- paste0(capt1, "\n", capt2)

p <- pdata |> 
  filter((geotype=="nation") |
           (geotype=="state" & geoid=="36")) |> 
  ggplot(aes(endyear, value, colour=geotypef)) +
  geom_line() +
  geom_point() +
  scale_colour_manual(values=c("blue", "darkgreen")) +
  scale_y_continuous(name="Percentage of household income",
                     limits=c(0, NA), 
                     breaks=seq(0, 1, 0.05),
                     labels=scales::percent_format(accuracy=1)) +
  scale_x_continuous(name=NULL, breaks=seq(2000, 2020, 2)) +
  labs(caption = capt) +
  ggtitle("Housing costs as percentage of household income, 2006 to 2021") +
  facet_wrap(~tenuref) +
  theme_bw() +
  caption_left  +
  theme(legend.position = 'bottom') +
  legend_notitle
p

```

Note: To get trends over time we have to use a slightly different data source. Elswhere in this document we use HUD CHAS data, an enhanced version of the American Community Survey that is based on the 5-year ACS 2015-2019. The graph above uses annual data from the ACS.
